---
title: "Visualizing Financial Data"
subtitle: "Analytics for Financial Data Science"
author: Mike Aguilar
format:
  html:
    toc: true
    toc-depth: 5
    toc-location: left
    code-fold: true
editor: visual
execute:
  warning: false
  echo: true
---

# Background

- An investment analyst monitors four assets and wants to compare their historical performance and risk.
- The practical goal is to develop a small interactive dashboard that lets a user select two assets and a date range.
- The dashboard will display:
  - an indexed-price chart for comparing performance; and
  - return histograms for comparing the distributions of daily returns.
- We will build the application progressively:
  1.  create static visualizations with `ggplot2`;
  2.  reorganize comparisons with `facet_wrap()`;
  3.  add interactivity with `plotly`; and
  4.  connect user inputs to multiple visualizations with Shiny.

# Housekeeping

Task: Reset the R session and load the packages used in this lesson.

```{r}
#| warning: false

cat("\014")  # clear the console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
suppressPackageStartupMessages(library(plotly))
```

# Data

## Application

The data represent four fictional exchange-traded funds. The series are simulated so that the lesson is reproducible and does not depend on an external data provider.

| Ticker | Asset           | Simulated Characteristics                 |
|--------|-----------------|-------------------------------------------|
| TECH   | Technology Fund | Higher growth and volatility              |
| BANK   | Banking Fund    | Moderate growth with a temporary drawdown |
| ENERGY | Energy Fund     | Cyclical and relatively volatile          |
| BOND   | Bond Fund       | Lower growth and volatility               |

The simulated differences are instructional features. They allow us to evaluate whether different visualizations reveal different aspects of the same data.

## Generate

Task: Load a reproducible sample of daily prices for four fictional assets. The data are simulated once by [`GenerateDataVizDemoData.R`](GenerateDataVizDemoData.R) and saved to `Data/DataVizDemoData.rds`, so this document and the companion Shiny app (`app.R`) share exactly the same observations. Weekdays are used as a simple approximation to trading days.

```{r}
data_path <- here("?", "?")

if (!file.exists(data_path)) {
  source(here("Topics", "VizForFDS", "GenerateDataVizDemoData.R"))
}

viz_demo_data <- readRDS(data_path)
asset_data <- viz_demo_data$asset_data
asset_specifications <- viz_demo_data$asset_specifications

head(asset_data) %>%
  mutate(
    price = round(price, 2),
    daily_return = round(daily_return, 4),
    indexed_price = round(indexed_price, 2)
  ) %>%
  kable(caption = "First six simulated asset-date observations")
```

## Data Dictionary

Task: Create a data dictionary for the analytical dataset.

```{r}
data_dictionary <- tribble(
  ~variable,        ~type,      ~description,
  "date",           "Date",     "Approximate trading date",
  "ticker",         "Character", "Short asset identifier",
  "asset_name",     "Character", "Descriptive asset name",
  "price",          "Numeric",   "Simulated closing price",
  "daily_return",   "Numeric",   "One-period simple return in decimal form",
  "indexed_price",  "Numeric",   "Price indexed to 100 on the asset's first date"
)

kable(
  data_dictionary,
  caption = "Data dictionary"
)
```

Q: What does one row of `asset_data` represent? Why is this long format useful for visualization?

A:

?

## Validate

Task: Verify that every ticker-date combination is unique and that the variables used in the lesson do not contain missing values.

```{r}
duplicate_observations <- asset_data %>%
  dplyr::count(ticker, date, name = "n") %>%
  filter(n > 1)

missing_summary <- asset_data %>%
  summarise(across(everything(), ~ sum(?(.x)))) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_values"
  )

if (nrow(duplicate_observations) > 0) {
  stop("Resolve duplicated ticker-date observations before continuing.")
}

if (any(missing_summary$missing_values > 0)) {
  stop("Resolve missing values before continuing.")
}

message("No duplicated ticker-date observations were found.")
missing_summary %>%
  kable(caption = "Missing-value summary")
```

# Static Visualization

## The Grammar of Graphics

The grammar of graphics treats a visualization as a collection of components that can be assembled and modified systematically.

``` r
ggplot(data, aes(x = ..., y = ...)) +
  geom_*() +
  scale_*() +
  labs() +
  theme_*()
```

| Component | Role | Example |
|------------------------|------------------------|------------------------|
| Data | Observations supplied to the plot | `asset_data` |
| Aesthetic mapping | Connects variables to visual properties | `aes(x = date, y = price)` |
| Geometry | Determines how observations are drawn | `geom_line()` |
| Scale | Controls how data values become visual values | `scale_y_continuous()` |
| Labels | Communicate titles, axes, legends, and sources | `labs()` |
| Theme | Controls non-data elements | `theme_minimal()` |

Q: Why is ggplot code connected with `+` signs?

A:
?

### One Asset: Basic Line Plot

Task: Filter the data to TECH and create a line plot of its price over time.

```{r}
tech_data <- asset_data %>%
  filter(?)

tech_basic_plot <- tech_data %>%
  ggplot(aes(x = ?, y = ?)) +
  geom_line()

tech_basic_plot
```

Q: What information is supplied by `ggplot(aes(x = date, y = price))`, and what information is supplied by `geom_line()`?

A:

?


Q: Why is a line plot usually more useful than an unconnected scatterplot for an ordered price series?

A:

?

### Aesthetic Mappings and Fixed Properties

Task: Make the TECH line dark blue and slightly thicker.

```{r}
tech_fixed_color_plot <- tech_data %>%
  ggplot(aes(?)) +
  ?(
    ? = "#00539B",
    ? = 0.9
  )

tech_fixed_color_plot
```

Q: Why are `color` and `linewidth` placed outside `aes()` in this plot?

A:

?

### Labels, Scales, and Themes

Task: Refine the TECH plot by adding informative labels, formatting the vertical axis as dollars, applying a minimal theme, and moving the title and source into a clear visual hierarchy.

```{r}
tech_polished_plot <- tech_data %>%
  ggplot(aes(x = date, y = price)) +
  geom_line(
    color = "#00539B",
    linewidth = 0.9
  ) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_dollar(accuracy = 1)
  ) +
  labs(
    ? = "Technology Fund Price",
    ? = "Simulated daily closing prices",
    x = NULL,
    ? = "Price",
    caption = "Source: Simulated data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "gray40")
  )

tech_polished_plot
```

Q: How do labels, scales, and themes differ?

A:
?

Q: Why might a polished chart remove minor grid lines but retain major grid lines?

A:
?

## Comparing Multiple Assets

### Multiple Lines: Raw Prices

Task: Create one plot containing the raw prices of all four assets. Map `ticker` to color.

```{r}
raw_price_plot <- asset_data %>%
  ggplot(
    aes(
      x = date,
      y = price,
      color = ?
    )
  ) +
  geom_line(linewidth = 0.75) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_dollar(accuracy = 1)
  ) +
  labs(
    title = "Asset Prices",
    subtitle = "Raw prices are not a common performance scale",
    x = NULL,
    y = "Price",
    color = "Ticker",
    caption = "Source: Simulated data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

raw_price_plot
```

Q: What causes ggplot to draw four lines instead of connecting all observations into one line?

A:

?

Q: Why is comparing raw price levels generally not the same as comparing investment performance?

A:
?

### Multiple Lines: Indexed Prices

For each asset, indexed price is:

$$
IndexedPrice_{i,t} = 100\left(\frac{Price_{i,t}}{Price_{i,0}}\right).
$$

Task: Plot all four assets using prices indexed to 100 on each asset's first observation.

```{r}
indexed_price_plot <- asset_data %>%
  ggplot(
    aes(
      x = ?,
      y = ?,
      color = ?
    )
  ) +
  geom_hline(
    yintercept = 100,
    color = "gray55",
    linewidth = 0.5,
    linetype = "dashed"
  ) +
  geom_line(linewidth = 0.8) +
  scale_x_date(
    date_breaks = "3 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1)
  ) +
  labs(
    title = "Indexed Asset Prices",
    subtitle = "Each asset equals 100 on its first observation",
    x = NULL,
    y = "Indexed Price",
    color = "Ticker",
    caption = "Source: Simulated data"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

indexed_price_plot
```

Q: How should we interpret an indexed price of 112? How should we interpret an indexed price of 94?

A:

?

Q: What analytical comparison becomes easier in the indexed-price plot? What information from the raw-price plot is no longer visible?

A:

?
## Faceted Comparisons

### Small Multiples

Task: Use `facet_wrap()` to place each asset's indexed price in a separate panel. Arrange the panels in two columns and retain a common vertical scale.

```{r}
faceted_price_plot <- asset_data %>%
  ggplot(aes(x = date, y = indexed_price)) +
  geom_hline(
    yintercept = 100,
    color = "gray55",
    linewidth = 0.45,
    linetype = "dashed"
  ) +
  geom_line(
    color = "#00539B",
    linewidth = 0.75
  ) +
  ?(
    vars(?),
    ncol = ?
  ) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b\n%Y",
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1)
  ) +
  labs(
    title = "Indexed Asset Prices by Ticker",
    subtitle = "Common axes preserve comparisons across panels",
    x = NULL,
    y = "Indexed Price",
    caption = "Source: Simulated data"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )

faceted_price_plot
```

Q: What are the respective advantages of the overlaid and faceted indexed-price plots?

A:
?

### Common and Free Scales

Task: Create a second faceted plot with a free vertical scale in each panel.

```{r}
faceted_free_y_plot <- faceted_price_plot +
  facet_wrap(
    vars(ticker),
    ncol = 2,
    scales = "?"
  ) +
  labs(
    title = "Indexed Asset Prices by Ticker",
    subtitle = "Free vertical scales emphasize within-asset variation"
  )

faceted_free_y_plot
```

Q: What does a free vertical scale make easier to see? What comparison does it make harder or potentially misleading?

A:

?

# Interactive Visualization

## Plotly

### Motivation

Q: What can an interactive chart provide that a static chart cannot?

A:
?

### Convert a ggplot

Task: Convert the indexed-price ggplot to an interactive Plotly figure.

```{r}
indexed_price_plotly_basic <- ?(
  indexed_price_plot,
  tooltip = c("x", "y", "colour")
) %>%
  plotly::layout(
    legend = list(orientation = "h")
  )

indexed_price_plotly_basic
```

Q: What is gained by passing an existing ggplot to `ggplotly()`? What control might eventually require native Plotly code?

A:

?

### Purposeful Tooltips

Task: Build the indexed-price chart with native Plotly syntax. Design a tooltip that reports the asset, date, indexed price, and cumulative return.

```{r}
plotly_tooltip_data <- asset_data %>%
  mutate(
    cumulative_return = indexed_price / 100 - 1,
    tooltip = paste0(
      "<b>", asset_name, "</b>",
      "<br>Date: ", format(date, "%B %d, %Y"),
      "<br>Indexed price: ", number(indexed_price, accuracy = 0.1),
      "<br>Cumulative return: ", percent(cumulative_return, accuracy = 0.1)
    )
  )

indexed_price_plotly <- ?(
  data = plotly_tooltip_data,
  x = ~date,
  y = ~indexed_price,
  color = ~ticker,
  type = "scatter",
  mode = "lines",
  text = ~tooltip,
  hoverinfo = "text"
) %>%
  plotly::layout(
    title = list(
      text = paste0(
        "Indexed Asset Prices",
        "<br><sup>Hover for exact values; select legend items to show or hide series</sup>"
      ),
      x = 0.02
    ),
    xaxis = list(title = ""),
    yaxis = list(title = "Indexed Price"),
    legend = list(
      title = list(text = "Ticker"),
      orientation = "h",
      x = 0,
      y = -0.18
    ),
    hovermode = "closest"
  )

indexed_price_plotly
```

Q: Why not place every available variable in the tooltip?

A:

?
### Date-Range Controls

Task: Add one-month, three-month, six-month, one-year, and full-sample buttons to the horizontal axis. Also add a range slider.

```{r}
indexed_price_plotly_controls <- indexed_price_plotly %>%
  plotly::layout(
    # The two-line title (title + hover subtitle) needs more room than the
    # default top margin, which otherwise crowds the range-selector buttons.
    # The added range slider also needs more room below the axis, so the
    # bottom margin grows and the legend moves further down to clear it.
    margin = list(t = 110, b = 100),
    legend = list(
      title = list(text = "Ticker"),
      orientation = "h",
      x = 0,
      y = -0.35
    ),
    xaxis = list(
      title = "",
      rangeselector = list(
        buttons = list(
          list(
            count = 1,
            label = "1m",
            step = "month",
            stepmode = "backward"
          ),
          list(
            count = 3,
            label = "3m",
            step = "month",
            stepmode = "backward"
          ),
          list(
            count = 6,
            label = "6m",
            step = "month",
            stepmode = "backward"
          ),
          list(
            count = 1,
            label = "?",
            step = "year",
            stepmode = "backward"
          ),
          list(
            label = "All",
            step = "all"
          )
        )
      ),
      rangeslider = list(visible = TRUE)
    )
  )

indexed_price_plotly_controls
```

Q: Does selecting a shorter Plotly date range change the indexed-price calculation?

A:

?

Q: How does clicking a Plotly legend entry differ from filtering the underlying data?

A:

?

## Shiny

### What Shiny Adds

Shiny connects three elements:

1.  **Inputs:** controls through which the user communicates a selection.
2.  **Reactive expressions:** calculations that rerun when their inputs change.
3.  **Outputs:** tables, plots, or other objects rebuilt from reactive results.

In our application, the user selects two assets and a date range. One reactive dataset filters the observations and re-indexes both selected assets to 100 on the first selected date. Both dashboard visualizations consume that same reactive dataset.

Q: What is the main difference between Plotly interactivity and Shiny reactivity in this lesson?

A:

?

Q: Why does the Shiny code below appear with `eval: false` instead of running inside this document?

A:
?
### Embedded Shiny Document or Separate `app.R`?

| Design | Advantages | Limitations |
|------------------------|------------------------|------------------------|
| Shiny embedded in a QMD | Integrates exposition, tasks, code, answers, and the live application; supports the progressive structure of this lesson; keeps the instructional artifact together | Requires a Shiny-capable R process; the rendered document is not a standalone static HTML file; a long lesson can mix instructional and application code |
| Separate `app.R` | Follows the conventional Shiny application structure; easier to expand into modules, supporting files, tests, and a larger deployment; separates the application from the lesson narrative | Requires students to move between the QMD and another file; duplicates or separates some context; weakens the one-document progression used here |

For this lesson, the dashboard code is still presented inline, with `eval: false`, so the reactive structure builds progressively alongside the narrative. It actually runs from a separate `app.R`, which keeps the instructional flow of an embedded document while avoiding the cost — and the project-preview fragility described above — of requiring a live Shiny session just to render the lesson. A larger dashboard, with modules, multiple pages, supporting assets, or independent deployment, would make a fully separate `app.R` design worthwhile regardless of this concern.

::: {.callout-important collapse="false" title="Execution Requirement"}
The code chunks in this section have `eval: false`: they are shown for exposition and do not run when this document is rendered. The live dashboard is the separate file `app.R`, located alongside this document. It loads the same simulated data (`Data/DataVizDemoData.rds`, produced by `GenerateDataVizDemoData.R`), so it stays in sync with the code shown here.

Launch the dashboard with:

``` r
shiny::runApp("app.R")
```

or open `app.R` in Positron/RStudio and use **Run App**.
:::

### User Interface

Task: Create inputs that allow the user to select two assets and a date range. Create placeholders for an indexed-price chart and faceted return histograms.

```{r}
#| echo: true
#| eval: false

fluidRow(
  column(
    width = 5,
    selectizeInput(
      inputId = "selected_assets",
      label = "Select exactly two assets:",
      choices = setNames(
        asset_specifications$ticker,
        paste0(
          asset_specifications$ticker,
          " — ",
          asset_specifications$asset_name
        )
      ),
      selected = c("TECH", "BANK"),
      multiple = TRUE,
      options = list(
        maxItems = 2,
        plugins = list("remove_button")
      )
    )
  ),
  column(
    width = 5,
    dateRangeInput(
      inputId = "selected_dates",
      label = "Select a date range:",
      start = min(asset_data$date),
      end = max(asset_data$date),
      min = min(asset_data$date),
      max = max(asset_data$date),
      format = "yyyy-mm-dd"
    )
  )
)

fluidRow(
  column(
    width = 12,
    plotlyOutput(
      outputId = "shiny_price_plot",
      height = "430px"
    )
  )
)

fluidRow(
  column(
    width = 12,
    plotlyOutput(
      outputId = "shiny_return_histograms",
      height = "390px"
    )
  )
)
```

Q: What roles do `inputId` and `outputId` play?

A:

?

### Reactive Data

Task: Create one reactive expression that validates the inputs, filters the selected observations, and re-indexes each asset to 100 on the first date in the selected range.

```{r}
#| eval: false

selected_asset_data <- reactive({
  validate(
    need(
      length(input$selected_assets) == 2,
      "Select exactly two assets."
    ),
    need(
      length(input$selected_dates) == 2,
      "Select a valid start and end date."
    )
  )

  filtered_data <- asset_data %>%
    filter(
      ticker %in% input$selected_assets,
      date >= input$selected_dates[1],
      date <= input$selected_dates[2]
    )

  validate(
    need(
      nrow(filtered_data) > 0,
      "The selected date range contains no observations."
    ),
    need(
      n_distinct(filtered_data$ticker) == 2,
      "The selected period must contain observations for both assets."
    )
  )

  filtered_data %>%
    group_by(ticker) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(
      selected_period_index = 100 * price / first(price),
      selected_period_return = selected_period_index / 100 - 1
    ) %>%
    ungroup()
})
```

Q: Why place the filtering and indexing logic in one reactive expression rather than repeat it inside both output definitions?

A:
?

Q: Why recompute `selected_period_index` rather than use the original `indexed_price` variable?

A:
?

### Reactive Indexed-Price Chart

Task: Render an interactive indexed-price chart from `selected_asset_data()`.

```{r}
#| eval: false

output$shiny_price_plot <- renderPlotly({
  plot_data <- selected_asset_data() %>%
    mutate(
      tooltip = paste0(
        "<b>", asset_name, "</b>",
        "<br>Date: ", format(date, "%B %d, %Y"),
        "<br>Selected-period index: ",
        number(selected_period_index, accuracy = 0.1),
        "<br>Selected-period return: ",
        percent(selected_period_return, accuracy = 0.1)
      )
    )

  plot_ly(
    data = plot_data,
    x = ~date,
    y = ~selected_period_index,
    color = ~ticker,
    type = "scatter",
    mode = "lines",
    text = ~tooltip,
    hoverinfo = "text"
  ) %>%
    plotly::layout(
      title = list(
        text = paste0(
          "Selected Assets: Indexed Prices",
          "<br><sup>Each asset equals 100 on its first selected date</sup>"
        ),
        x = 0.02
      ),
      xaxis = list(title = ""),
      yaxis = list(title = "Indexed Price"),
      legend = list(
        title = list(text = "Ticker"),
        orientation = "h",
        x = 0,
        y = -0.18
      ),
      hovermode = "closest"
    )
})
```

### Reactive Return Histograms

Task: Render faceted histograms of daily returns for the same selected observations. Use common bins and axes so that the return distributions remain comparable.

```{r}
#| eval: false

output$shiny_return_histograms <- renderPlotly({
  histogram_data <- selected_asset_data()

  return_histogram_plot <- histogram_data %>%
    ggplot(aes(x = daily_return, fill = ticker)) +
    geom_histogram(
      binwidth = 0.005,
      boundary = 0,
      color = "white",
      linewidth = 0.25,
      alpha = 0.85
    ) +
    geom_vline(
      xintercept = 0,
      color = "gray35",
      linewidth = 0.5,
      linetype = "dashed"
    ) +
    facet_wrap(vars(ticker), ncol = 2) +
    scale_x_continuous(
      labels = label_percent(accuracy = 1)
    ) +
    labs(
      title = "Selected Assets: Daily Return Distributions",
      subtitle = "Common bins and axes support cross-asset comparison",
      x = "Daily Return",
      y = "Number of Observations"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "none",
      plot.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )

  ggplotly(
    return_histogram_plot,
    tooltip = c("x", "count")
  )
})
```

Q: Why do the histograms use `daily_return` rather than `selected_period_index`?

A:

?

Q: Why use common histogram bins and axes across the two facets?

A:

?

### Dashboard Interpretation

Task: Use the dashboard to compare TECH and BOND over the full sample and then over a shorter period. Evaluate both cumulative performance and the dispersion of daily returns.

Q: Why might changing the selected date range alter both the apparent performance ranking and the return histograms?

A:

?

Q: What evidence would support the statement that one selected asset was riskier than another? What limitations remain?

A:
?

# Visualization Design Principles

Q: What general principles from this lesson extend beyond financial time series?

A:

?

# Common Errors

Q: What common implementation and interpretation errors should we avoid?

A:
?