---
title: "Financial Data Fundamentals"
subtitle: "Cleaning & Engineering"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
  enabled: false
---

# Housekeeping

```{r}
#| warning: FALSE

cat("\014")  # clear console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Cleaning

## Missing

Missing values (NA) occur when an observation is unavailable. Before deciding how to handle them, determine:

How many are missing? Where are they located (beginning, middle, or end)? Why are they missing?

Common remedies include:

- Delete observations
- Forward fill
- Linear interpolation

Task: create a tibble called prices with two columns. Date: starting 1Jan2026, formatted as a date, and progressing for 7 days. Price with values 100, 101,missing, missing, 105, missing, 107.

```{r}
prices <- tibble(
  Date = seq(as.Date("2026-01-01"), by = "day", length.out = 7),
  Price = c(100, 101, NA, NA, 105, NA, 107)
)
prices
```

Task: sum over the Price colum using is.na

```{r}
?
```

Q: What is the meaning of this output?

A: ?

Task: Run which(is.na(prices\$Price))

```{r}
?
```

Q: Interpret the output.

A: ?

Task: Drop the rows with missing values

```{r}
?
```

Task: use na.approx to fill in the missing values

```{r}
prices %>%
  mutate(Price = zoo::?)
```

Q: What is na.approxy doing?

A: ?

Task: use fill from tidyr with direction down to overcome the missing values.

```{r}
prices %>%
  tidyr::?(?, .direction = "?")
```

Q: What is fill doing? Why might that be useful?

A:

?

## Duplicates

Duplicate observations often arise after downloading data from multiple sources or merging datasets.

Duplicates should usually be identified using the variables that uniquely identify an observation.

Task: Create a simple tibble of Date and Price. Let's repeat the 2026-01-02 record

```{r}
prices <- tibble(
  Date = as.Date(c(
    "2026-01-01",
    "2026-01-02",
    "2026-01-02",
    "2026-01-03")),
  Price = c(100,101,101,102)
)

prices
```

Task: use "duplicated()" from dyplyr

```{r}
prices %>%
  ?
```

Q: What is duplicated() doing for us?

A: ?

Task: Use "distinct()" from dyplyr

```{r}
prices %>%
?
```

Q: What is distinct() doing for us?

A: ?

## Invalid

Some observations are impossible or inconsistent.

Q: Can you think of a few examples within our domain of Financial Data Science?

A:

?

Task: Create a simple tibble with a negative price

```{r}
prices <- tibble(
  Date = seq(as.Date("2026-01-01"), by="day", length.out=5),
  Price = c(100,101,-5,103,104)
)

prices
```

Task: filter on negative prices via dyplyr

```{r}
prices %>%
?
```

Q: What's the relevance in our context?

A: ?

Task: filter on "valid" prices via dyplyr

```{r}
prices %>%
?
```

Q: Why might this approach be troublesome?

A:

?

## Sorting

Many financial calculations assume observations are in chronological order.

Always verify the data are sorted before computing

Q: What is the best practice when sorting? oldest to youngest or vice versa?

A: ?

Task: Create a small tibble of prices that is out of order

```{r}
prices <- tibble(
  Date = as.Date(c(
    "2026-01-03",
    "2026-01-01",
    "2026-01-02")),
  Price = c(102,100,101)
)

prices
```

Task: arrange by date using dyplyr

```{r}
prices %>%
?
```

## Filtering

Filtering selects only the observations that satisfy specified conditions.

Common examples in our financial data science context include

- One ticker
- One year
- Positive returns
- Trading days after a given date

Task: Create a simple tibble starting at 2026-01-01. Sequence for 5 consecutive day and populate with prices 100, 101, 102, 103,104

```{r}
prices <- tibble(
  Date = seq(as.Date("2026-01-01"), by="day", length.out=5),
  Price = c(100,101,102,103,104)
)

prices
```

Task: use dyplyr filter to grab only the observations after 2Jan

```{r}
prices %>%
?
```

Task: Create a simple tibble with AAPL and MSFT. We observe each for a single point in 2025 and 2026. The return for AAPL is .1 and .08 for 2025, 2026, respectively, and .15, .12 for MSFT.

```{r}
stocks <- tibble(
  Ticker = c("AAPL","AAPL","MSFT","MSFT"),
  Year = c(2025,2026,2025,2026),
  Return = c(.10,.08,.15,.12)
)
```

Task: Use dyplyr filter to extract only AAPL in 2026

```{r}
stocks %>%
?
```

# Working with Dates

## Formatting

Sometimes, the databases from which we extract financial data don't have dates that are formatted in a manner that is easy for analysis.

Let's explore a few data vendors, their date formats, and correct as needed.

### Yahoo finance

Task: mimic a small sample of data from Yahoo Finance

```{r}
prices <- tibble(
  Date = as.Date(c(
    "2026-01-02",
    "2026-01-05",
    "2026-01-06")),
  Price = c(100,101,103)
)

prices
```

Task: Use class to determine the date format

```{r}
?
```

Q: Any concerns?

A: ?

### Ken French

Task: Create a small tibble to mimic the date formats from Ken French's data repo

```{r}
factors <- tibble(
  Date = c(202510, 202511, 202512),
  Mkt_RF = c(1.2, -0.8, 2.5)
)

factors
```

Task: Use class to determine the date format

```{r}
?
```

Q: Any concerns?

A:

Task: Use dyplyr to create a new Date object via the ymd function

```{r}
factors <- factors %>%
  mutate(
    Date = lubridate::?(paste0(Date, "?"))
  )

factors
```

### FRED

Task: Create a small tiblle to mimic dates from FRED

```{r}
cpi <- tibble(
  Date = as.Date(c(
    "2025-10-01",
    "2025-11-01",
    "2025-12-01")),
  CPI = c(322.5,323.0,323.6)
)

cpi
```

```{r}
?
```

Q: Any concerns?

A: ?

## Aggregation

A very common engineering task when dealing with time series data, is aggregating; e.g. days to weeks; months to years.

Task: Create a simple tibble of sales data starting at 2026-01-05. Use "seq" to carry those dates 14 total calendar days. Add a price column with values 100, 101, ...113.

```{r}
prices <- tibble(
  Date = seq(as.Date("2026-01-05"), by = "day", length.out = 14),
  Price = c(100:113)
)

prices
```

Task: Add the day of the week to the prices object

```{r}
prices<- prices %>%
  mutate(DayOfWeek = ?)
prices
```

Task: Add a column that marks the date of the first day of each week in which that date appears.

```{r}
prices <- prices %>%
  mutate(
    WeekFirstDay = lubridate::?(
      Date,
      unit = "week",
      week_start = 1
    )
  )

prices
```

Task: Add a column that marks the price of the first day of each week in which that date appears.

```{r}
prices <- prices %>%
  group_by(WeekFirstDay) %>%
  mutate(
    WeekFirstPrice = ?
  ) %>%
  ungroup()

prices
```

Task: Add a column that marks the date of the last day of each week in which that date appears.

```{r}
prices <- prices %>%
  group_by(WeekFirstDay) %>%
  mutate(
    WeekLastDay = ?
  ) %>%
  ungroup()

prices
```

Task: Add a column that marks the price of the last day of each week in which that date appears.

```{r}
prices <- prices %>%
  group_by(WeekFirstDay) %>%
  mutate(
    WeekLastPrice = ?
  ) %>%
  ungroup()

prices
```

Task: Add a column that marks the average of the prices for the week in which that date appears.

```{r}
prices <- prices %>%
  group_by(WeekFirstDay) %>%
  mutate(
    WeekAvgPrice = ?
  ) %>%
  ungroup()

prices
```

# Reshaping

## Wide vs Long

Task: create a wide tibble with Date (2026-01-31, 2026-02-28, 2026-03-31). Then add 3 columns of prices. For AAPL (190,195,198), MSFT (420, 425, 430), and NVDA (140, 145, 150)

```{r}
stocks_wide <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-02-28",
    "2026-03-31")),
  AAPL = c(190,195,198),
  MSFT = c(420,425,430),
  NVDA = c(140,145,150)
)

stocks_wide
```

Q: Suppose we want to convert the wide dataframe to long. Describe that long object; how many columns? how many rows? what are the columns? what are the rows?

A: ?

Task: Convert the wide dataframe to long

```{r}
stocks_long <- ? %>%
  ?(
    cols = -?,
    names_to = "?",
    values_to = "?"
  ) 

stocks_long
```

Task: Rearrange that long object by ticker and within each ticker arrange by date in ascending order

```{r}
stocks_long %>%
    ?

```

Task: Reshape the long back into wide

```{r}
stocks_long %>%
  ?(
    names_from = ?,
    values_from = ?
  )
```

## Selecting

We often want to examine a subset of our fields/columns. Within dyply we can use "select". Moreover, we often want to restrict our attention to a subset of values. Within dyplyr we can use "filter".

Task: From the wide dataframe, create a subset of stocks_wide that has only Date, APPL, and MSFT

```{r}
stocks_wide %>%
?
```

Task: From the long dataframe, create a subset of stocks_long that focuses on APPL and MSFT

```{r}
stocks_long %>%
?
```

## Joining

We will often need to join/merge datasets and data objects. For instance, suppose we wanted to explore the relationship between inflation and asset returns. We need to create an object that contains both of these values. Importantly, the rows need to be aligned such that the dates align. There are several ways to join. in this small exercise, we will examine left_join.

Task: Create a small tibble for dates 2026-01-31, 2026-02-28, 2026-03-31, and returns .025, -.012, .041.

```{r}
returns <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-02-28",
    "2026-03-31")),
  Return = c(0.025,-0.012,0.041)
)

returns
```

Task: Create a small tibble for dates 2026-01-31, 2026-03-31 and inflation .003, .004

```{r}
inflation <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-03-31")),
  Inflation = c(0.003,0.004)
)

inflation
```

Task: create an object "combine" that is a left_join between returns and inflation by Date.

```{r}
combined <- ?
combined
```

Task: recreate the combined object but reverse the order of inflation and returns in the left_join

```{r}
combined <- ?
combined
```

Q: Why are they different?

A: ?

# Returns Engineering

Within financial data science one of the seminal target variables is return. Although it may seem simple, the concept and calculation of returns requires great care.

## Simple vs Log

Task: Create a simple tibble of monthly returns starting from 2026-01-31 and progressing for a total of 5 months. Add prices of 98, 104, 102, 108, 111.

```{r}
prices <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-02-28",
    "2026-03-31",
    "2026-04-30",
    "2026-05-31"
  )),
  Price = c(98, 104, 102, 108, 111)
)

prices
```

Task: create an object called "returns" that has Gross Simple Returns, Net Simple Returns, and Net Log Returns

```{r}
returns <- prices %>%
  mutate(
    GrossSimpleReturn = ?,
    NetSimpleReturn   = ?,
    NetLogReturn   = ?
  )

returns
```

## Multi horizon returns

Our databases sometimes have returns, not prices. Moreover, returns are often used in modeling and comparison. These stylized facts leave return as the core data object. That return, however, may not be observed over the horizon of interest. For example, we may have daily returns, but want monthly.

Task: Compute rolling 2 month Gross Simple Return, and rolling 2 month new Simple Return

```{r}
returns %>%
  mutate(
    GrossSimple2 = ?,
    NetSimple2   = ?
  )

```

Task: Compute rolling 2 month Net Log Return

```{r}
returns %>%
  mutate(
    NetLog2 = ?
  )
```

## Annualization

Q: What is the motivation for annualization in our typical financial data science setting?

A:

Task: Suppose the daily_return was 0.0008. Convert to an annual return. You may assume 252 trading days per year.

```{r}
daily_return <- 0.0008
?
```

Task: Suppose the quarter_return was .05. Convert to an annual return.

```{r}
quarter_return <- 0.05

?
```

Task: Suppose the total return over a 4 year period was 40%. Convert to an annual return.

```{r}
total_return <- 0.40
years <- 4

?
```

## Indexing

Q: Explain what (re)indexing means and it's potential use within our financial data science context.

A: ?

Task: create an index for the price object, starting from 1

```{r}
prices %>%
  mutate(
    Index = ?
  )
```

Q: What does the index mean in our case?

A: ?

## Lags

Lags are commonplace in financial data science. In fact, we've already used a lag operator in order to generate returns. Let's expand slightly.

Task: Add columns to the "prices" object that contains 1,2,3 lags

```{r}
prices<-prices %>%
  mutate(
    Lag1 = ?,
    Lag2 = ?,
    Lag3 = ?
  )
  prices
```

Task: Add columns to the "prices" object that contains 1,2,3 leads

```{r}
prices<-prices %>%
  mutate(
    Lead1 = ?,
    Lead2 = ?,
    Lead3 = ?
  )
  prices
```

## Inflation adjustment

Q: Can you think of a use case for inflation adjustment within our financial data science context?

A:

?

Task: Create a simple tibble for the months of Jan, Feb, Mar, Apr 2026 with Nominal returns = .03, .02, -.01, .04 and month over month Inflation = .004, .003, .002, .005

```{r}
returns <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-02-28",
    "2026-03-31",
    "2026-04-30"
  )),
  Nominal = c(0.03, 0.02, -0.01, 0.04),
  Inflation = c(0.004, 0.003, 0.002, 0.005)
)

returns
```

Task: Add a column "Real" using the exact Fisher formula

```{r}
returns<-returns %>%
  mutate(
    Real = ?
  )
```

Task: Add a columns "ApproxReal" which shows the real return using the Fisher approximation, and ApproxError which is the Approx Real minus Real

```{r}
returns %>%
  mutate(
    ApproxReal = ?,
    ApproxError = ?
  )
```

Q: In what circumstance might the Approximation Error grow?

A: ?

Task: Create a simple tibble for Jan, Feb, Mar, Apr 2026 with Prices 100, 103, 105, 108 and CPI index level 120,120.4, 120.8, 121.2. Let's assume that the CPI is indexed at 2020-01-31 = 100.

```{r}
prices <- tibble(
  Date = as.Date(c(
    "2026-01-31",
    "2026-02-28",
    "2026-03-31",
    "2026-04-30"
  )),
  Price = c(100, 103, 105, 108),
  CPI = c(120.0, 120.4, 120.8, 121.2)
)
prices
```

Task: Add the Real Price to the prices object

```{r}
CPI_base <- 100

prices <- prices %>%
  mutate(
    RealPrice = ?
  )
  prices
```

Q: What is the meaning of the 2026-04-30 Real Price?

A: ?