---
title: "Financial Data Fundamentals"
subtitle: "Systematic Factors"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
  enabled: false # use to false when executing "skeletons" to avoid issues with incomplete code chunks
---

# Housekeeping

```{r}
#| warning: FALSE

cat("\014")  # clear console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Data Overview

## Data Type

Systematic factor datasets summarize the returns to broad sources of risk that are believed to explain differences in expected asset returns. Rather than representing the performance of a single security, each factor is itself a time series constructed from diversified portfolios.

Common equity factors include

- Market Excess Return (Mkt-RF)
- Size (SMB)
- Value (HML)
- Profitability (RMW)
- Investment (CMA)
- Momentum (UMD / MOM)

Factors are typically reported at

- Daily
- Weekly
- Monthly
- Annual

frequencies.

------------------------------------------------------------------------

## Why It Matters

Systematic factors are among the most widely used datasets in empirical finance.

Applications include

- Measuring portfolio performance
- Estimating abnormal returns (alpha)
- Risk attribution
- Asset pricing research
- Portfolio construction
- Factor investing
- Performance benchmarking
- Academic replication studies

Nearly every empirical asset pricing paper begins by downloading one or more factor datasets.

------------------------------------------------------------------------

## Data Structure

Unlike OHLCV data, factor datasets are already transformed into returns.

A typical monthly Fama-French dataset contains

| Variable | Description                   |
|----------|-------------------------------|
| Date     | Observation month             |
| Mkt-RF   | Market excess return          |
| SMB      | Small minus Big               |
| HML      | High minus Low book-to-market |
| RF       | Risk-free rate                |

Later datasets may additionally include

| Variable | Description                              |
|----------|------------------------------------------|
| RMW      | Robust minus Weak profitability          |
| CMA      | Conservative minus Aggressive investment |
| MOM      | Momentum factor                          |

Most datasets report returns in **percent**, not decimal form.

Unlike stock prices, factor datasets generally

- contain no missing trading days
- require little cleaning
- have consistent frequencies
- are already adjusted for dividends and stock splits

------------------------------------------------------------------------

## Providers

### Academic Data Libraries

- **Kenneth French Data Library (Dartmouth)** — industry standard source for Fama-French factors and portfolio sorts
- **AQR Data Library** — quality, value, momentum, carry, defensive factors
- **Hou-Xue-Zhang q-Factor Library** — q-factor model datasets

### Commercial Vendors

- WRDS
- Bloomberg
- FactSet
- LSEG / Refinitiv
- MSCI Barra

### Government Sources

While governments do not publish factor returns directly, several underlying inputs originate from

- CRSP
- Compustat
- U.S. Treasury
- Federal Reserve

------------------------------------------------------------------------

## Access Methods

Common acquisition methods include

- Website downloads (CSV or ZIP)
- REST APIs
- R packages
  - `tidyfinance`
  - `tidyverse`
  - `download.file()`
  - `httr`
- Python packages
  - `pandas-datareader`
  - `requests`
  - `famafrench`

Unlike FRED or Alpha Vantage, the Kenneth French Data Library does not currently provide a formal REST API. Instead, users download ZIP archives directly from the server.

------------------------------------------------------------------------

## Data Challenges

Common issues include

- Returns are reported in **percent** rather than decimal form.
- Dates are encoded as integers (e.g. 202501).
- Multiple datasets use different frequencies.
- Files often contain documentation above and below the actual data.
- Missing values often indicate the end of the dataset rather than missing observations.
- Different datasets use different naming conventions (MOM vs UMD).
- Some factors are excess returns while others are raw returns.

Students should always inspect

- units
- frequency
- date format
- construction methodology
- documentation

before beginning analysis.

------------------------------------------------------------------------

## Factor Zoo

The Fama-French three-factor model is only one member of a much larger family of systematic factors.

Examples include

| Family        | Representative Factors                     |
|---------------|--------------------------------------------|
| CAPM          | Market                                     |
| Fama-French 3 | Market, SMB, HML                           |
| Carhart 4     | FF3 + Momentum                             |
| Fama-French 5 | FF3 + RMW + CMA                            |
| Fama-French 6 | FF5 + Momentum                             |
| q-Factors     | Market, Size, Investment, Profitability    |
| AQR           | Value, Momentum, Quality, Carry, Defensive |

Throughout the remainder of the course we will primarily use the Fama-French datasets because they are freely available, well documented, widely accepted in both academia and industry, and easily reproducible.

## Example Analyses

Common applications include

- Estimate CAPM beta
- Estimate FF3 alpha
- Estimate FF5 exposures
- Build factor mimicking portfolios
- Compare active managers against factor benchmarks
- Compute rolling factor loadings
- Study historical factor premiums
- Test new asset pricing models

# Data Acquisition Exercises

## Web

Tasks

- Visit Ken French's Data Library https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/data_library.html
- Download the FF3 monthly data
- Unzip the file into the Data directory
- Open in Excel. Scroll down the columns. Don't skip; scroll. What do you see as you get near the end of the daily data?
- Load that csv file into R

Task: Load the csv such that we only import the monthly data and we ignore the first 3 rows of descriptive text.

```{r}
Factors_Web <- read_csv(here::here("?","?"), skip = 4) %>%
  rename(Date = 1) %>%
  filter(grepl("^\\d{6}$", Date)) %>%
  mutate(Date = as.integer(Date))

```

Task: Reload the data. Call it "raw". Don't filter that object.

```{r}
raw <- read_csv(
  here::here("Data","F-F_Research_Data_Factors.csv"),
  skip = 4,
  show_col_types = FALSE
) %>%
  rename(Date = 1)
```

Task: Now filter on those that are NOT equal to the 6 digit Dates

```{r}
raw %>%
  filter(!grepl("^\\d{6}$", Date)) %>%
  dplyr::select(Date) %>%
  head(10)
```

Q: What is the relevance?

A: ?

## API

Let's now automate the whole process of downloading the zip and extracting the csv.

Task: Set file name and path on the French library site

```{r}
French_ftp <- "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/"
French_core_filename<-c("F-F_Research_Data_Factors")
French_zip_filename<-paste0(French_core_filename,"_CSV.zip")
French_url<- paste0(French_ftp,French_zip_filename)
French_url
```

Task: Establish where you want to unzip the file and define the name you want to assign

```{r}
zipfile <- ?("Data", "FF_Factors.zip")
```

Task: Download the zip file

```{r}
download.file(?, zipfile, mode = "wb")
```

Task: Unzip the file

```{r}
ff_factors_raw_data <- ?(?, exdir = here("Data"))
```

Task: Establish name of csv you want to load

```{r}
French_csv_file<-paste0(?,".csv")
```

Task: Read the contents of the csv and pull the monthly observations only

```{r}
Factors_API <- read_csv(here::here("Data",?), skip = 4) %>%
  rename(Date = 1) %>%
  filter(grepl("^\\d{6}$", Date)) %>%
  mutate(Date = as.integer(Date))
```

## Compare

Task: Display the first 3 rows of the Web and API datasets

```{r}
head(Factors_Web,3)
head(Factors_API,3)
```

# Reconstructing FF3

- Pull the 6 portfolios formed on Size and Book to Market
- Reconstruct the SMB
- Reconstruct the HML
- What is the market return? (Not the market minus risk free)

## Download Portfolios

Goal: Download the 6 portfolios formed by Size and BtM

Task:

```{r}
French_core_filename<-c("6_Portfolios_2x3")
French_zip_filename<-paste0(?,"_CSV.zip")
French_url<- paste0(French_ftp,?)
French_url

```

Task: Establish where you want to unzip the file and define the name you want to assign

```{r}
zipfile <- here("?", "FF_6Portfolios.zip")
```

Task: Download the zip file

```{r}
download.file(?, zipfile, mode = "wb")
```

Task: Unzip the file

```{r}
six_six_btm_portfolios <- ?(zipfile, exdir = here("Data"))
```

Task: Define the file name that you want to read

```{r}
French_csv_file<-paste0(French_core_filename,"?")
French_csv_file
```

Task: Load the csv

```{r}
SixPort <- read_csv(here::here("Data",?), skip = 15) %>%
  rename(Date = 1) 
```

Task: Keep only the first chunk of data since there are several variants in the csv.

```{r}
SixPort <- SixPort %>%
  slice(
    1:(which(is.na(Date))[1] - 1)
  ) %>%
  mutate(
    Date = as.integer(Date),
    across(-Date, as.numeric)
  )

SixPort<-na.omit(SixPort)
```

## Reconstruct Size and Value

Task: Reconstruct SMB

```{r}
SMB_reconstructed <-
?
```

Task: Reconstruct HML

```{r}
HML_reconstructed <-
?
```

Task: Compare the portfolios from the French library to our reconstruction

```{r}
comparison <- tibble(

    Date = Factors_API$Date,

    SMB = Factors_API$SMB,

    SMB_Reconstructed = SMB_reconstructed,

    SMB_Error = Factors_API$SMB - SMB_reconstructed,

    HML = Factors_API$HML,

    HML_Reconstructed = HML_reconstructed,

    HML_Error = Factors_API$HML - HML_reconstructed
)

summary(comparison$SMB_Error)

summary(comparison$HML_Error)
```