---
title: "OHLCV"
subtitle: "Data For Financial Data Science"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
engine: knitr
execute:
  warning: false
  echo: true
---

```{r}
#| echo: true
cat("\014")  # clear console
rm(list = ls())  # clear the workspace

if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Data Acquisition Exercises

## Excel

- Open MSFT Excel
  - Navigate to Data -\> Data Types. If you see "Stocks" you should be able to proceed. Usually available for MSFT 365 accounts on Windows and MAC.
- Type CAT in cell A1
- With cursor on CAT hit the Stocks button
- Select CATERPILLAR in the Data Selector bar on the right
- Hover over sell A1 and hit the Insert Data box.
- Click previous close

## Web via Investing.com

- Visit https://www.investing.com/
- Create a free account
- Search for CAT
- Click Historical Data
- Daily
  - Note that the default frequency is Daily. If not, switch to daily.
  - Alter the date range to 1Dec2025 through 31Dec2025
  - Download that data
- Monthly
  - Switch the Time Frame to Monthly
  - Set the range from 1Dec2025 through 31Dec2025
  - Download that data
- Excel
  - Create an Excel file called CAT within the Data folder of the class repo cloned to your machine
  - Create a sheet called Daily and paste the daily CAT prices starting in cell A1
  - Create a sheet called Comparison, and paste in the daily prices and nearby paste the monthly prices
- Comparison
  - Reconcile "Price" from Monthly with "Price" from Daily
  - Reconcile "Open" from Monthly with "Open" from Daily
  - Reconcile "High" from Monthly with "High" from Daily
  - Reconcile "Low" from Monthly with "Low" from Daily
  - Solutions
    - Note: 12/1/2025 represents the month of December, not the 1st day
    - Price(Dec) = Price(12/31/2025)
    - Open(Dec) = Open(12/1/2025)
    - High(Dec) = Highest daily High
    - Low(Dec) = Lowest daily Low
  - Load the data from the Daily sheet into R

Task: Set path to Data folder and CAT xlsx file

```{r}
path_cat <- here("Data", "CAT.xlsx")
```

Task: use read_excel to pull the Daily sheet

```{r}
CAT_Excel <- read_excel(path_cat, sheet = "Daily")
```

## API via Yahoo Finance

Goal: Acquire a historical time series of stock prices via Yahoo Finance API.

Architecture:

- Ensure quantmod is invoked
- Set date range
- Set ticker
- Call API

Task: Use requireNamespace to ensure quantmod is installed. Use isNamespaceload to ensure quantmod library is invoked. These should both return TRUE since we ran our housekeeping earlier.

```{r}
print(requireNamespace("quantmod", quietly = TRUE))
print(isNamespaceLoaded("quantmod")
)
```

Task: Create an empty list objecct to store our parameters

```{r}
params <- list()
```

Task: Set a date range to start 1Jan2025 and end 31Dec2025. Note that dates are typically entered yyymmdd. Add start and end to params

```{r}
params$start <- "2025-01-01"
params$end <- "2025-12-31"
```

Task: Set the ticker to CAT and add to params

```{r}
ticker = "CAT"
params$ticker<-ticker
```

Task: Call to API via getSymbols

```{r}
getSymbols(params$ticker,from=params$start,to=params$end,src="yahoo")
```

Task: Display the first 10 rows. Also display the last 10 rows.

```{r}
head(CAT)
tail(CAT)
```

Task: display the data for 12/30/2025 from the Excel upload

```{r}
target_date <- ymd("2025-12-30")
CAT_Excel %>% filter(Date == target_date)

```

Task: Display the data for 12/30/2025 from the Yahoo Finance API dataset

```{r}
CAT[target_date]
```

"What would the price history look like if dividends and splits had never created mechanical jumps?"

```{r}
CAT$AdjFactor <- CAT$CAT.Adjusted / CAT$CAT.Close

head(CAT[, c("CAT.Close", "CAT.Adjusted", "AdjFactor")])
```

# Exploring Adjusted Close

Q: What is the meaning of "Adjusted Close"?

A: historical price series that removes the effects of dividends and stock splits, allowing returns to be calculated consistently over time.

## Case A

Assume the stock drops from 100 on 1/16/26 to 99 on 1/20/26, and that a $\$1$ dividend was received on 1/20/26.

Q: What is the adjusted closing price on each day? Explain.

A:

| Date |  Close | ΔClose |  Div | Wealth | ΔWealth | Adj Close | ΔAdj |
|:-----|-------:|-------:|-----:|-------:|--------:|----------:|-----:|
| 1/16 | 100.00 |     -- | 0.00 | 100.00 |      -- |     99.00 |   -- |
| 1/20 |  99.00 |  -1.00 | 1.00 | 100.00 |    0.00 |     99.00 | 0.00 |

- The stock price fell from 100.00 to 99.00.
- It appears the investor lost 1.00.
- However, the investor also received a 1.00 dividend.
- Therefore, the investor's total wealth is $99.00 + 1.00 = 100.00$
- So no wealth was lost
- Although the Close price fell, the investor's total wealth did not change. The drop in the stock price simply reflects that 1.00 of value was transferred from the company to shareholders as a cash dividend.
- The Adjusted Close removes this mechanical dividend effect so that returns reflect changes in total investment value rather than changes in the quoted stock price alone.

## Case B

Assume a stock drops from 100 on 1/16/26 to 97.5 on 1/20/26, and that a $\$1$ dividend was received on 1/20/26.

Q: What is the adjusted closing price on each day? Explain.

A:

| Date |  Close | ΔClose |  Div | Wealth | ΔWealth | Adj Close |  ΔAdj |
|:-----|-------:|-------:|-----:|-------:|--------:|----------:|------:|
| 1/16 | 100.00 |     -- | 0.00 | 100.00 |      -- |     99.00 |    -- |
| 1/20 |  97.50 |  -2.50 | 1.00 |  98.50 |   -1.50 |     97.50 | -1.50 |

- The stock price fell from 100.00 to 97.50, a decline of 2.50.
- At first glance, it appears the investor lost 2.50.
- However, the investor also received a 1.00 dividend.
- Therefore, the investor's total wealth is 97.50 + 1.00 = 98.50
- which is a loss of only 98.50 - 100.00 = -1.50
- Notice the Dividend-Adjusted Close column.
  - The January 16 closing price is adjusted downward by the 1.00 dividend to 99.00.
  - The January 20 closing price is unchanged at 97.50 because it is already trading after the dividend has been paid.
  - The return computed from the Dividend-Adjusted Close series is $\frac{97.50-99.00}{99.00}=-1.52\%$, which closely matches the investor's true loss of 1.50 on a 100 investment.
- This illustrates the purpose of an adjusted price series: it removes the portion of the price change that is simply due to the dividend, leaving a return that better reflects the change in the value of the investment.

## CAT's Dividend

From CAT's website https://investors.caterpillar.com/stock-info/dividend-history/default.aspx we know there was a \$1.510 per share dividend on 1/20/206. Let's use that to understand the meaning of Adjusted Closing Price in practice.

Task: Grab CAT from 12/31/2025 through 1/31/2026

```{r}
params$start = "2025-12-31"
params$end = "2026-01-31"
getSymbols(params$ticker,from=params$start,to=params$end,src="yahoo")
```

Task: Convert to a data frame and add a date

```{r}
CAT_df <- as.data.frame(CAT)
CAT_df <- tibble::rownames_to_column(CAT_df, var = "Date")
CAT_df$Date <- as.Date(CAT_df$Date)
```

Task: Create a Dividend column of 1.51 for the appropriate date

```{r}

CAT_df <- CAT_df %>%
  mutate(
    Dividend = if_else(Date == as.Date("2026-01-20"), 1.51, 0)
  )
```

Task: Create a column that computes the Delta in the Close price

```{r}
CAT_df <- CAT_df %>%
    mutate( 
      CloseDelta = CAT.Close - lag(CAT.Close)
    )
```

Task: Create a column for the Wealth, assuming the investor owns one share

```{r}
CAT_df <- CAT_df %>%
  mutate(
    Wealth = CAT.Close + Dividend
  )
```

Task: Create a column for the Delta in Wealth

```{r}
CAT_df <- CAT_df %>%
  mutate(
    WealthDelta = Wealth - lag(CAT.Close)
  )
```

Task: Create a column for the Delta in the Adjusted Close

```{r}
CAT_df <- CAT_df %>%
  mutate(
    AdjDelta = CAT.Adjusted - lag(CAT.Adjusted)
  )
```

Task: Filter on the dates 1/16 and 1/20 and display the Date, Close, Close Delta, Wealth, Wealth Delta, Adjusted, and Adjusted Delta

```{r}
CAT_window <- CAT_df %>%
  filter(Date %in% as.Date(c("2026-01-16", "2026-01-20")))%>%
  dplyr::select(Date,CAT.Close,CloseDelta,Wealth,WealthDelta,CAT.Adjusted,AdjDelta)

CAT_window
```

Q: Explain the Adjusted Closing logic in this case.

A:

- On January 20, 2026, CAT closed at 629.00, a decline of 17.89 from the previous close of 646.89. -However, CAT also paid a 1.51 per share dividend that day. An investor who owned the stock before the market opened on January 20 received this cash payment.
- Therefore, the investor's wealth at the end of the day was 629+1.51=630.51, not just 629
- The investor's wealth therefore fell by 630.51-646.89=-16.38, rather than the apparent 17.89 decline suggested by the closing prices alone.
- Adjusted Close is designed to account for these dividend payments so that returns reflect changes in investor wealth, not simply changes in the quoted stock price.
- To accomplish this, Yahoo Finance does not adjust the January 20 closing price. Instead, it retroactively adjusts the prices before January 20 so that returns computed from the Adjusted Close series are comparable across the dividend event.
- Notice that the Adjusted Close is not exactly equal to "Close + Dividend." Yahoo Finance applies cumulative historical adjustment factors for all dividends and stock splits. The important idea is not the exact algorithm, but that Adjusted Close produces a price series suitable for computing total returns

## Loading Multiple Tickers

Goal: Load multiple tickers and place the adjusted closing price into a wide dataframe.

Task: Set the dates from 31Dec2025 through 1Jan2026

```{r}
params$start <- "2025-12-31"
params$end   <- "2026-01-31"
```

Task: Create a character vector for AAPL, CAT, AMZN

```{r}
tickers <- c("AAPL", "CAT","AMZN")
```

Task: Download all series

```{r}
getSymbols(
  Symbols = tickers,
  from = params$start,
  to = params$end,
  src = "yahoo"
)
```

Task: Merge adjusted closes into one xts object

```{r}
AdjPrices <- do.call(
  merge,
  lapply(tickers, function(x) Ad(get(x)))
)
```

Task: Rename columns with the ticker names

```{r}
colnames(AdjPrices) <- tickers

AdjPrices
```