---
title: "OHLCV"
subtitle: "Data For Financial Data Science"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
engine: knitr
execute:
  warning: false
  echo: true
  enabled: false # use to false when executing "skeletons" to avoid issues with incomplete code chunks
---



```{r}
#| echo: true
cat("\014")  # clear console
rm(list = ls())  # clear the workspace

if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

# Data Acquisition Exercises

## Excel

- Open MSFT Excel
  - Navigate to Data -\> Data Types. If you see "Stocks" you should be able to proceed. Usually available for MSFT 365 accounts on Windows and MAC.
- Type CAT in cell A1
- With cursor on CAT hit the Stocks button
- Select CATERPILLAR in the Data Selector bar on the right
- Hover over sell A1 and hit the Insert Data box.
- Click previous close

## Web via Investing.com

- Visit https://www.investing.com/
- Create a free account
- Search for CAT
- Click Historical Data
- Daily
  - Note that the default frequency is Daily. If not, switch to daily.
  - Alter the date range to 1Dec2025 through 31Dec2025
  - Download that data
- Monthly
  - Switch the Time Frame to Monthly
  - Set the range from 1Dec2025 through 31Dec2025
  - Download that data
- Excel
  - Create an Excel file called CAT
  - Create a sheet called Daily and paste the daily CAT prices starting in cell A1
  - Create a sheet called Comparison, and paste in the daily prices and nearby paste the monthly prices
- Comparison
  - Reconcile "Price" from Monthly with "Price" from Daily
  - Reconcile "Open" from Monthly with "Open" from Daily
  - Reconcile "High" from Monthly with "High" from Daily
  - Reconcile "Low" from Monthly with "Low" from Daily
  - Solutions??? 
  - Load the data from the Daily sheet into R


Task: Set path to Data folder and CAT xlsx file

```{r}
path_cat <- here("?", "?")
```

Task: use read_excel to pull the Daily sheet

```{r}
CAT_Excel <- ?(path_cat, sheet = "?")
```

## API via Yahoo Finance

Goal: Acquire a historical time series of stock prices via Yahoo Finance API.

Architecture:

- Ensure quantmod is invoked
- Set date range
- Set ticker
- Call API

Task: Use requireNamespace to ensure quantmod is installed. Use isNamespaceload to ensure quantmod library is invoked. These should both return TRUE since we ran our housekeeping earlier.

```{r}
print(requireNamespace("quantmod", quietly = TRUE))
print(isNamespaceLoaded("quantmod")
)
```

Task: Create an empty list objecct to store our parameters

```{r}
params <- ?
```

Task: Set a date range to start 1Jan2025 and end 31Dec2025. Note that dates are typically entered yyymmdd. Add start and end to params

```{r}
params$start <- "2025-01-01"
params$? <- "2025-12-31"
```

Task: Set the ticker to CAT and add to params

```{r}
ticker = "?"
params$ticker<-?
```

Task: Call to API via getSymbols

```{r}
?(params$ticker,from=params$start,to=params$end,src="yahoo")
```

Task: Display the first 10 rows. Also display the last 10 rows.

```{r}
head(?)
tail(?)
```

Task: display the data for 12/30/2025 from the Excel upload

```{r}
target_date <- ymd("2025-12-30")
CAT_Excel %>% filter(Date ??)

```

Task: Display the data for 12/30/2025 from the Yahoo Finance API dataset

```{r}
CAT?
```

"What would the price history look like if dividends and splits had never created mechanical jumps?"

```{r}
CAT$AdjFactor <- CAT$CAT.Adjusted / CAT$CAT.Close

head(CAT[, c("CAT.Close", "CAT.Adjusted", "AdjFactor")])
```

# Exploring Adjusted Close

Q: What is the meaning of "Adjusted Close"?

A: ?

## Case A

Assume the stock drops from 100 on 1/16/26 to 99 on 1/20/26, and that a 1 dividend was received on 1/20/26.

Q: What is the adjusted closing price on each day? Explain.

A:

?

## Case B

Assume a stock drops from 100 on 1/16/26 to 97.5 on 1/20/26, and that a 1 dividend was received on 1/20/26.

Q: What is the adjusted closing price on each day? Explain.

A:

?

## CAT's Dividend

From CAT's website https://investors.caterpillar.com/stock-info/dividend-history/default.aspx we know there was a \$1.510 per share dividend on 1/20/206. Let's use that to understand the meaning of Adjusted Closing Price in practice.

Task: Grab CAT from 12/31/2025 through 1/31/2026

```{r}
params$start = "2025-12-31"
params$end = "?"
getSymbols(params$?,from=params$start,to=params$end,src="yahoo")
```

Task: Convert to a data frame and add a date

```{r}
CAT_df <- as.data.frame(CAT)
CAT_df <- tibble::rownames_to_column(CAT_df, var = "Date")
CAT_df$Date <- as.Date(CAT_df$Date)
```

Task: Create a Dividend column of 1.51 for the appropriate date

```{r}

CAT_df <- CAT_df %>%
  ?(
    Dividend = if_else(Date == as.Date("2026-01-20"), ?, 0)
  )
```

Task: Create a column that computes the Delta in the Close price

```{r}
CAT_df <- CAT_df %>%
    mutate( 
      CloseDelta = CAT.Close - ?(CAT.Close)
    )
```

Task: Create a column for the Wealth, assuming the investor owns one share

```{r}
CAT_df <- CAT_df %>%
  mutate(
    ? = CAT.Close + Dividend
  )
```

Task: Create a column for the Delta in Wealth

```{r}
CAT_df <- CAT_df %>%
  mutate(
    WealthDelta = Wealth - ?
  )
```

Task: Create a column for the Delta in the Adjusted Close

```{r}
CAT_df <- CAT_df %>%
  mutate(
    AdjDelta = CAT.Adjusted - ?
  )
```

Task: Filter on the dates 1/16 and 1/20 and display the Date, Close, Close Delta, Wealth, Wealth Delta, Adjusted, and Adjusted Delta

```{r}
CAT_window <- CAT_df %>%
  filter(Date %in% as.Date(c("2026-01-16", "2026-01-20")))%>%
  dplyr::select(Date,CAT.Close,CloseDelta,Wealth,WealthDelta,CAT.Adjusted,AdjDelta)

CAT_window
```

Q: Explain the Adjusted Closing logic in this case.

A:

?

## Loading Multiple Tickers

Goal: Load multiple tickers and place the adjusted closing price into a wide dataframe.

Task: Set the dates from 31Dec2025 through 1Jan2026

```{r}
params$start <- "2025-12-31"
params$end   <- "2026-01-31"
```

Task: Create a character vector for AAPL, CAT, AMZN

```{r}
tickers <- ?
```

Task: Download all series

```{r}
getSymbols(
  Symbols = ?,
  from = params$start,
  to = params$end,
  src = "yahoo"
)
```

Task: Merge adjusted closes into one xts object

```{r}
AdjPrices <- do.call(
  merge,
  lapply(tickers, function(x) Ad(get(x)))
)
```

Task: Rename columns with the ticker names

```{r}
?(AdjPrices) <- tickers

AdjPrices
```