---
title: "EDA for Financial Data Science"
author: Mike Aguilar
format: html
editor: visual
toc: true
toc-depth: 5
toc-location: left
embed-resources: true
execute:
  warning: false
  echo: true
  enabled: false
---

# Housekeeping

```{r}
#| warning: FALSE

cat("\014")  # clear console
rm(list = ls())  # clear the workspace
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
suppressPackageStartupMessages(library(here))
source(here("Supporting", "ProjectSetup.R"))
```

Goal: Summarize the behavior of the simple daily net returns for IBM, FORD, CAT, AAPL, and DIS from the beginning of 2020 through the end of 2025

# Data

## Data Acquisition

Task: Create a parameter list of the tickers and dates

```{r}
params = list(
  tickers = c("IBM", "F", "CAT", "AAPL", "DIS"), 
  startdate = "2019-12-20", 
  enddate = "2025-12-31"
)
```

Q: If we want 2020 through 2025, why are we grabbing data from 2019.

A: ? 

Task: Load the tickers with the getSymbols function

```{r}
getSymbols(
  ?,
  src = "yahoo",
  from = ?,
  to = ?,
  index.class = "POSIXct"
)
```

Task: Create a wide data object containing the adjusted closing prices for all of the assets.

```{r}
AdjPrices_Wide <- do.call(
  merge,
  lapply(params$tickers, function(x) Ad(get(x)))
)
```

Task: Add the date to that wide object

```{r}
AdjPrices_Wide <- AdjPrices_Wide %>%
  as_tibble(rownames = "date") %>%
  mutate(date = as.Date(?))
```

Task: Transpose the wide object to long format

```{r}
AdjPrices_Long <- AdjPrices_Wide %>%
  ?(-date, names_to = "ticker", values_to = "adj_close")
```

Task: Rename the columns so that they conform to the tickers and does not have the .Adjusted suffix

```{r}
AdjPrices_Long <- AdjPrices_Long %>%
  ?(ticker = str_remove(ticker, "\\.Adjusted$"))
```


We will use this long data object throughout the rest of the exercise, unless otherwise stated. 

## Cleaning

Task: Peak at the head and tail of the long data

```{r}
head(AdjPrices_Long)
tail(AdjPrices_Long)
```

Task: Count missing

```{r}
AdjPrices_Long %>%
  ?(across(everything(), ~sum(?(.))))
```

Task: Count negative prices

```{r}
AdjPrices_Long %>%
  group_by(ticker) %>%
  ?(n_negative = sum(adj_close ?))
```

Q: Any concerns?

A: ?

## Engineering

Task: Generate simple daily net returns

```{r}
Returns_Long <- AdjPrices_Long %>%
  arrange(ticker, date) %>%
  group_by(?) %>%
  mutate(return = ?) %>%
  ungroup()
```

Q: Why is the group_by vital?

A: ?

Task: Verify the returns avoid this trap by filtering on the last days of AAPL and the first days of CAT. 

```{r}
tickers_sorted <- sort(unique(Returns_Long$ticker))
current <- "AAPL"
next_ticker <- tickers_sorted[which(tickers_sorted == current) + 1]

bind_rows(
  Returns_Long %>% filter(ticker == current) %>% tail(),
  Returns_Long %>% filter(ticker == next_ticker) %>% head()
)
```


Q: How do we know we didn't fall into the trap? 

A: ?

Task: Trim the returns object so that it only covers the date range of interest. 

```{r}
Returns_Long<-Returns_Long %>%
  filter(?)
min(Returns_Long$date)
```
Task: Clean the missing observations resulting from this engineering.  Use na.omit. 

```{r}
Returns_Long <- ?
```

# Exploration

## Univariate Summary

Task: Run the summary function on the return column from Returns_Long.

```{r}
summary(Returns_Long$?)
```

Q: What's potentially misleading with this approach?

A: ?

Task: Create a custom dyplr call that generates mean, std dev, skew, kurt, min, q10, q25, q50, q75, q90, max. Apply to each asset.

```{r}
Returns_Long %>% 
  dplyr::select(ticker, return) %>%
  group_by(ticker) %>%
  summarise(across(where(is.numeric), 
  .fns = list(
    min = ~round(min(., na.rm = TRUE), 2),
    median = ~round(median(., na.rm = TRUE), 2),
    mean = ~round(mean(., na.rm = TRUE), 3),
    stdev = ~round(sd(., na.rm = TRUE), 3),
    skew = ~round(skewness(.,na.rm=TRUE),2),
    kurt = ~round(kurtosis(.,na.rm=TRUE),2), # Not excess
    q10 = ~round(quantile(., 0.10, na.rm = TRUE), 2),
    q25 = ~round(quantile(., 0.25, na.rm = TRUE), 2),
    q75 = ~round(quantile(., 0.75, na.rm = TRUE), 2),
    q90 = ~round(?(., ?, na.rm = TRUE), 2),
    max = ~round(?(., na.rm = TRUE), 2)
  ),
  .names = "{fn}"
  )) 
```

Q: Does the data follow a normal distribution?

A: ?

Task: Focus on the first asset from the ticker list within your parameter settings. Generate a frequency histogram of returns.

```{r}
first_ticker<-params$tickers[1]
temp<-Returns_Long %>%
  filter(ticker == ?) %>%
  select(?)
hist(
  temp$return,
  main = first_ticker, 
  xlab = "Simple Net Daily Return (.05=5%)"
)
```

Task: Using that same asset, generate a relative frequency histogram with an overlay of a Normal distribution. Also add a subtitle indicating the start and end dates for returns. 

```{r}
mu <- mean(temp$return)
sigma <- sd(temp$return)
first_day <- Returns_Long$date[?]
last_day <- Returns_Long$date[?]
subtitle_text <- paste0(?," through ",?)

ggplot(temp, aes(x = return)) +
  ?(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "white") +
  stat_function(fun = dnorm, args = list(mean = mu, sd = sigma), color = "firebrick", linewidth = 1) +
  labs(
    title = paste("Relative Frequency Histogram of", first_ticker, "Daily Returns"),
    subtitle = subtitle_text,
    x = "Daily Return (0.05 = 5%)",
    y = "Density (10 = 10%)"
  ) +
  theme_minimal()
```

Q: Does the data follow a normal distribution?

A: ?

Task: Using that same asset, generate a QQ Plot

```{r}
?(temp$return, main = paste("Normal Q-Q Plot:", first_ticker, "Daily Returns"))
qqline(temp$return, col = "firebrick")
```

Q: Interpret the QQ Plot

A: ?

Task: Generate a single figure with violin plots of the densities for each asset side by side.

```{r}
ggplot(Returns_Long, aes(x = ticker, y = return)) +
  ?(fill = "steelblue") +
  labs(
    title = "Distribution of Daily Returns by Asset",
    x = "Ticker",
    y = "Simple Daily Return (.1=10%)"
  ) +
  theme_minimal()
```

Q: Which asset looks most extreme in this violin plot?

A: ?

Task: Build a table of the raw return quantiles (q1, q10, q25, q50, q75, q90, q99) for each ticker, with a Normal(0,1) reference row for comparison.

```{r}
probs <- c(0.01, 0.10, 0.25, 0.50, 0.75, 0.90, 0.99)
q_names <- paste0("q", probs * 100)

raw_quantiles <- Returns_Long %>%
  group_by(ticker) %>%
  reframe(as_tibble_row(setNames(?(return, probs), q_names)))

normal_quantiles_raw <- as_tibble_row(setNames(?(probs), q_names)) %>%
  mutate(ticker = "Normal", .before = 1)

bind_rows(normal_quantiles_raw, raw_quantiles)
```

Q: Does this raw-scale table support the "F is most extreme" impression from the violin plot?

A:?

Task: Standardize each asset relative to its own history (z-score), then regenerate the violin plot on the standardized scale.

```{r}
Returns_Long_z <- Returns_Long %>%
  group_by(ticker) %>%
  mutate(z = ?) %>%
  ungroup()

ggplot(Returns_Long_z, aes(x = ticker, y = z)) +
  geom_violin(fill = "steelblue") +
  labs(
    title = "Distribution of Standardized (Z-Score) Daily Returns by Asset",
    x = "Ticker",
    y = "Z-Score"
  ) +
  theme_minimal()
```

Task: Compute the q1, q10, q25, q50, q75, q90, q99 of each asset's z score. Display in a table where these quantiles are the columns, and the rows pertain to the Normal distribution followed by each ticker.

```{r}
ticker_quantiles <- Returns_Long_z %>%
  group_by(ticker) %>%
  reframe(as_tibble_row(setNames(quantile(?, probs), q_names)))

normal_quantiles <- as_tibble_row(setNames(?(probs), q_names)) %>%
  mutate(ticker = "Normal", .before = 1)

bind_rows(normal_quantiles, ticker_quantiles)
```

Q: Now that we've standardized, is F still the most extreme?

A: ? 

## Evolution over time

Task: Focus on the first ticker again. Plot the adjusted closing price using the basic "plot" function.

```{r}
temp <- Returns_Long %>%
  filter(ticker == first_ticker) %>%
  select(date, adj_close)

plot(?, ?, type = "l",
     xlab = "", ylab = "Adjusted Close",
     main = paste(first_ticker, "Closing Price"))
```

Task: Using that same asset, plot the closing price for the last 6mths as a candlestick using the chart_Series function.

```{r}
?(x = last(get(first_ticker), "6 months"),
             type = "?", name = first_ticker)
```

Task: Using that same asset, create a drawdown chart using charts.PerformanceSummary.

```{r}
?(na.omit(Return.calculate(Cl(get(first_ticker)), method = "discrete")))
```

Task: Using that same asset, compute 20 day rolling mean returns. Also compute 250 day rolling mean returns. Create a plot overlaying these series.

```{r}
roll_df <- Returns_Long %>%
  filter(ticker == ?) %>%
  arrange(?) %>%
  mutate(
    roll_20  = zoo::rollmean(return, k = ?, fill = NA, align = "right"),
    roll_250 = zoo::rollmean(return, k = ?, fill = NA, align = "right")
  )

ggplot(roll_df, aes(x = date)) +
  geom_line(aes(y = roll_20, color = "20-day")) +
  geom_line(aes(y = roll_250, color = "250-day")) +
  labs(
    title = paste(first_ticker, "Rolling Mean Return"),
    x = "Date", y = "Rolling Mean Return", color = "Window"
  ) +
  theme_minimal()
```

Task: Using that same asset, compute 20 day rolling mean of returns SQUARED. Also compute 250 day rolling mean returns. Create a plot overlaying these series.

```{r}
roll_sq_df <- Returns_Long %>%
  filter(ticker == first_ticker) %>%
  arrange(date) %>%
  mutate(
    roll_sq_20  = zoo::rollmean(?, k = 20, fill = NA, align = "right"),
    roll_sq_250 = zoo::rollmean(?, k = 250, fill = NA, align = "right")
  )

ggplot(roll_sq_df, aes(x = date)) +
  geom_line(aes(y = roll_sq_20, color = "20-day")) +
  geom_line(aes(y = roll_sq_250, color = "250-day")) +
  labs(
    title = paste(first_ticker, "Rolling Mean of Squared Return"),
    subtitle = "Proxy for time-varying volatility",
    x = "", y = "Rolling Mean Squared Return", color = "Window"
  ) +
  theme_minimal()
```

Task: Using that same asset, compute 20 day rolling standard deviation of returns. Also compute 250 day rolling mean returns. Create a plot overlaying these series.

```{r}
roll_sd_df <- Returns_Long %>%
  filter(ticker == first_ticker) %>%
  arrange(date) %>%
  mutate(
    roll_sd_20  = zoo::rollapply(return, width = 20, FUN = ?, fill = NA, align = "right"),
    roll_sd_250 = zoo::rollapply(return, width = 250, FUN = ?, fill = NA, align = "right")
  )

ggplot(roll_sd_df, aes(x = date)) +
  geom_line(aes(y = roll_sd_20, color = "20-day")) +
  geom_line(aes(y = roll_sd_250, color = "250-day")) +
  labs(
    title = paste(first_ticker, "Rolling Standard Deviation of Return"),
    subtitle = "Proxy for time-varying volatility",
    x = "", y = "Rolling Standard Deviation", color = "Window"
  ) +
  theme_minimal()
```

Task: Using that same asset, compute 20 day rolling skewness returns. Also compute 250 day rolling mean returns. Create a plot overlaying these series.

```{r}
roll_skew_df <- Returns_Long %>%
  filter(ticker == first_ticker) %>%
  arrange(date) %>%
  mutate(
    roll_skew_20  = zoo::rollapply(return, width = 20, FUN = ?, fill = NA, align = "right"),
    roll_skew_250 = zoo::rollapply(return, width = 250, FUN = ?, fill = NA, align = "right")
  )

ggplot(roll_skew_df, aes(x = date)) +
  geom_line(aes(y = roll_skew_20, color = "20-day")) +
  geom_line(aes(y = roll_skew_250, color = "250-day")) +
  labs(
    title = paste(first_ticker, "Rolling Skewness of Return"),
    x = "", y = "Rolling Skewness", color = "Window"
  ) +
  theme_minimal()
```

Task: Using that same asset, compute 20 day rolling kurtosis returns. Also compute 250 day rolling mean returns. Create a plot overlaying these series.

```{r}
roll_kurt_df <- Returns_Long %>%
  filter(ticker == first_ticker) %>%
  arrange(date) %>%
  mutate(
    roll_kurt_20  = zoo::rollapply(return, width = 20, FUN = ?, fill = NA, align = "right"),
    roll_kurt_250 = zoo::rollapply(return, width = 250, FUN = ?, fill = NA, align = "right")
  )

ggplot(roll_kurt_df, aes(x = date)) +
  geom_line(aes(y = roll_kurt_20, color = "20-day")) +
  geom_line(aes(y = roll_kurt_250, color = "250-day")) +
  labs(
    title = paste(first_ticker, "Rolling Kurtosis of Return"),
    x = "", y = "Rolling Kurtosis", color = "Window"
  ) +
  theme_minimal()
```

Task: Create a plot that indexes the adjusted closing price to \$100 for each asset at the beginning of the sample and then tracks the evolution of that \$100 over time.

```{r}
indexed_df <- AdjPrices_Long %>%
  arrange(ticker, date) %>%
  group_by(ticker) %>%
  mutate(indexed_price = 1?) %>%
  ungroup()

ggplot(indexed_df, aes(x = date, y = indexed_price, color = ticker)) +
  geom_line() +
  labs(
    title = "Growth of $100 Invested, by Asset",
    x = "", y = "Value of $100 Investment", color = "Ticker"
  ) +
  theme_minimal()
```

## Comovement

Task: Pivot returns to wide format for matrix-based comovement analysis.

```{r}
Returns_Wide <- Returns_Long %>%
  select(date, ticker, return) %>%
  ?(names_from = ticker, values_from = return) %>%
  arrange(date)
```

Task: Compute the covariance matrix of returns

```{r}
cov_matrix <- Returns_Wide %>%
  select(-date) %>%
  ?()
round(cov_matrix, 6)
```

Task: Compute the correlation matrix of returns

```{r}
cor_matrix <- Returns_Wide %>%
  select(-date) %>%
  ?()
round(cor_matrix, 3)
```

Task: Visualize the comovement with pairs

```{r}
Returns_Wide %>%
  select(-?) %>%
  pairs(main = "Pairwise Scatterplots of Daily Returns")
```

Task: Visualization the correlation with chart.Correlation

```{r}
Returns_Wide %>%
  select(-date) %>%
  ?(histogram = TRUE)
```

Q: What do the red stars indicate? Interpret within our case. 

A:  ?

Task: Compute 250 day rolling correlation between the first and second tickers in the ticker list. Plot this relationship.

```{r}
second_ticker <- params$tickers[2]

roll_cor_df <- Returns_Wide %>%
  arrange(date) %>%
  mutate(
    roll_cor_250 = zoo::rollapply(
      cbind(.data[[first_ticker]], .data[[second_ticker]]),
      width = 250,
      FUN = function(x) ?(x[, 1], x[, 2]),
      by.column = FALSE,
      fill = NA,
      align = "right"
    )
  )

ggplot(roll_cor_df, aes(x = date, y = roll_cor_250)) +
  geom_line() +
  labs(
    title = paste("250-Day Rolling Correlation:", first_ticker, "vs", second_ticker),
    x = "", y = "Rolling Correlation"
  ) +
  theme_minimal()
```

Q: How stable is this correlation over time?

A: ?

## Temporal Dependence

Task: Compute the ACF for each asset's returns. Let's use ggACF applied to Returns_Wide wrapped in a function. 

```{r}
acf_plots <- purrr::map(params$tickers, function(tk) { 
  forecast::?(?[[tk]]) +
    ggplot2::labs(title = tk) 
})

patchwork::wrap_plots(acf_plots, ncol = 3)
```

Q: Is there significant autocorrelation in daily returns?

A: ?

Task: Pick the first ticker. Compute 5 lags of that asset's returns.

```{r}
lag_df <- Returns_Wide %>%
  arrange(date) %>%
  transmute(
    date,
    lag0 = .data[[first_ticker]],
    lag1 = lag(.data[[first_ticker]], ?),
    lag2 = lag(.data[[first_ticker]], ?),
    lag3 = lag(.data[[first_ticker]], ?),
    lag4 = lag(.data[[first_ticker]], ?),
    lag5 = lag(.data[[first_ticker]], ?)
  ) %>%
  na.omit()
```

Task: Visualize the correlation among those lags using ggpairs 

```{r}
ggpairs(
  ? %>% select(-date),
  upper = list(continuous = wrap("cor", size = 6)),
  diag  = list(continuous = wrap("densityDiag")),
  lower = list(continuous = wrap("points"))
)

```

Q: Interpret? 

A: ?

Task: Compute the ACF for each asset's squared returns. We can do this several ways.  In this exercise let's use ggAcf wrapped in a function. 

```{r}
sq_acf_plots <- purrr::map(params$tickers, function(tk) {
  ?(Returns_Wide[[tk]]^2) + labs(title = tk)
})

wrap_plots(sq_acf_plots, ncol = 3)
```

Q: How does this compare to the ACF of raw returns?

A: ?

Task: Use the same asset as above. Compute 5 lags of that asset's squared returns.

```{r}
sq_lag_df <- Returns_Wide %>%
  arrange(date) %>%
  transmute(
    date,
    lag0 = .data[[first_ticker]]?,
    lag1 = lag(.data[[first_ticker]]?, 1),
    lag2 = lag(.data[[first_ticker]]?, 2),
    lag3 = lag(.data[[first_ticker]]?, 3),
    lag4 = lag(.data[[first_ticker]]?, 4),
    lag5 = lag(.data[[first_ticker]]?, 5)
  ) %>%
  na.omit()
```

Task: Visualize the correlation among those lags. Use ggpairs. 

```{r}
ggpairs(
  ? %>% select(-?),
  upper = list(continuous = wrap("cor", size = 6)),
  diag  = list(continuous = wrap("densityDiag")),
  lower = list(continuous = wrap("points"))
)

```

Q: What is the implication of these results? 

A: ?

Task: Pick the first and second tickers Compute the CCF.

```{r}
?(Returns_Wide[[first_ticker]], Returns_Wide[[second_ticker]],
    main = paste("CCF:", first_ticker, "vs", second_ticker))
```

Q: What does the CCF show?

A: ?

Task: Lag the second asset 1,2,3,4,5 periods. Visualize the correlations between the first asset and the second asset's returns and it's lags.

```{r}
ccf_lag_df <- Returns_Wide %>%
  arrange(date) %>%
  transmute(
    date,
    asset1 = .data[[first_ticker]],
    lag0 = .data[[second_ticker]],
    lag1 = lag(.data[[second_ticker]], 1),
    lag2 = lag(.data[[second_ticker]], 2),
    lag3 = lag(.data[[second_ticker]], 3),
    lag4 = lag(.data[[second_ticker]], 4),
    lag5 = lag(.data[[second_ticker]], 5)
  ) %>%
  na.omit()

cor_vec <- sapply(ccf_lag_df %>% select(starts_with("lag")),
                   function(x) ?(ccf_lag_df$asset1, x))

cor_df <- tibble(lag = 0:5, correlation = cor_vec)

ggplot(cor_df, aes(x = lag, y = correlation)) +
  geom_col(fill = "steelblue") +
  labs(
    title = paste0("Correlation of ", first_ticker, "'s Return with ", second_ticker, "'s Lagged Return"),
    x = paste(second_ticker, "Lag (days)"), y = "Correlation"
  ) +
  theme_minimal()
```

Q: Does lagging the second asset's return reveal any predictive relationship?

A: 